---
title: "Week 3 HW"
author: "Ashley James"
date: "10/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#load packages

```{r}
library(tidyverse)
library(janitor)
library(skimr)

```
#import dataset
#this is a dataset from the NYC Department of Heath and human services on blood lead levels for children under 6. I manually cleaned the variable names because they very very long and detailed. 
```{r}
leadclean<- read_csv("leadclean.csv")
```

#use the filter and selcect functions to select only observations in Manhattan 
```{r}
leadclean %>% 
  filter(geo_area_name == "Manhattan") %>% 
  select(geo_area_name)
```
#Use filter to isoltae only the four boroughs in geo_area_name. Then group by geo_area_name. Then summarize by the mean bll5, which is the number of children under 6 with blood lead levels of 5 or greater. 
```{r}
avg_bll5_by_geo<- leadclean %>% 
  filter(geo_area_name %in% c("Manhattan" , "Brooklyn", "Queens", "Bronx", "Staten Island")) %>% 
  group_by(geo_area_name) %>% 
  summarise (meanbll5 = mean(bll5, na.rm = TRUE))
```