---
title: "Week 7 HW"
author: "Ashley James"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
The Dataset that I have been working with is already in long format. Tehrefore, I will take another dataset that is long format from the internet. I will use a dataset on child mortality from www.childmortality.org. 
#load packages 

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(readxl)
```
#import excel data 
#clean names and rename datatset clean

```{r}
child_mortality<-read_excel (path = "/Users/ashleyjames/Desktop/learning markdown/RatesDeaths_AllIndicators.xlsx" , sheet = "Rates and Deaths")
child_mortality_clean<-clean_names(child_mortality)
view(child_mortality_clean)
```
#use select to only include variables that relate to under 5 mornality (u5mr), country name, and uncertainty bounds
#then use pivot_longer to transform data from wide to long 
#then replace NAs with 0 
#name new dataset child_mortality_tidy and view 


```{r}
child_mortality_tidy<-child_mortality_clean %>% 
  select(country_name, uncertainty_bounds, contains("U5MR")) %>%
   pivot_longer(cols = u5mr_1950 : u5mr_2015, 
                names_to = "type_year", 
                values_to = "count") %>% 
  mutate(count = replace_na(count, 0)) 
View(child_mortality_tidy)

```



